### 2022-02-06 preprint.tex
Generalizing Scala's Singleton Types

In this chapter, we report on our attempt to generalize Scala's singleton types to support type-level programming, as well as a lightweight form a dependently-typed programming.
Unlike proof assistants, we do not aim to use types as a general-purpose logic, which would favor designs ensuring totality of functions through termination checks.
Instead, our focus is on improving type safety by increasing the expressive power of the type system.

TODO: reformulate
Our system generalizes the traditional notion of singleton type: when the lifted term $t$ contains a non-deterministic choice, the resulting type $\singleton{t}$ denotes the set of values that $t$ could possibly reduce to.
As a result, our type system is capable of type computations by manipulating types which are based on terms, but can nonetheless contain more than a single value.
In combination with subtyping, this allows us to seamlessly integrate with impure, or imprecisely-typed programs.

With these goals in mind, we extended Scala's type system to lift programs and partially evaluate them as part of type-checking.
Users can manipulate dependent types either directly, using new syntactic forms in Scala's type language, or have them inferred automatically using a new language keyword.
This effectively allows users to execute programs involving functions and pattern matching at the type level.

@@

\begin{itemize}
\item
  We begin by motivating why type-level programming in desirable and how one might use our Scala extension to improve type safety (\Cref{sec:motivating-example}). Our example demonstrates how to design a strongly-typed API in a functional style accessible to programmers.
\item
  We describe how we extended Scala with a generalization of singleton types.
  We prototyped our type system on top of Dotty, the reference Scala~3 compiler.
  This practical introduction would be of interest to any Scala programmer willing to learn how to use our system,
  as well as programming language designers interested dependent types.
\item
  We show a concrete use-case of our system by implementing a strongly-typed wrapper for Apache Spark~\citep{zaharia2016apache} (\Cref{sec:use-case}).
  Thanks to our generalized singleton types, we can statically ensure the type safety of database operations such as join and filter.
  We compare our implementation with an equivalent implicit-based one and show remarkable compilation time savings.
\end{itemize}

@@

For instance, consider the following list and its singleton type: |Cons(Nil, Nil): { Cons(Nil, Nil) }|.
In our system, this term may also be typed less precisely, such as:

>>>>

In this chapter, we report on our attempt to generalize Scala's singleton types to support type-level programming, as well as a lightweight form of dependently-typed programming.
Unlike proof assistants, we do not aim to use types as a general-purpose logic, which would favor designs ensuring the totality of functions through termination checks.
Instead, our focus is on improving type safety by increasing the expressive power of the type system.

With these goals in mind, we extended Scala's type system to lift programs to the type level and partially evaluate them as part of type-checking.
Users can manipulate those types either directly, using new syntactic forms in Scala's type language, or have them inferred automatically using a new language keyword.
This effectively allows users to execute programs involving functions and pattern matching at the type level.

@@

\begin{itemize}
\item
  We begin by motivating why type-level programming is desirable and how one might use our Scala extension to improve type safety (\Cref{sec:motivating-example}). Our example demonstrates how to design a strongly-typed API in a functional style accessible to programmers.
\item
  We describe how we extended Scala with a generalization of singleton types (\Cref{sec:extending-scala}).
  We prototyped our type system on top of Dotty, the reference Scala~3 compiler.
  This practical introduction would be of interest to any Scala programmer willing to learn how to use our system, as well as programming language designers interested in dependent types.
\item
  We show a concrete use-case of our system by implementing a strongly-typed wrapper for Apache Spark~\citep{zaharia2016apache} (\Cref{sec:use-case}).
  Thanks to our generalized singleton types, we can statically ensure the type safety of database operations such as join and filter.
  We compare our implementation with an equivalent implicit-based one and show remarkable compilation time savings.
\end{itemize}

Our original presentation of generalized singleton types contains a formalization \citep[Section 3 and 4]{schmid2020coming}, which is out of the scope of this chapter.

@@

For instance, consider the following list and its singleton type:

\begin{itemize}
  \item |Cons(Nil, Nil): { Cons(Nil, Nil) }|
\end{itemize}

In our system, this term may also be typed less precisely, such as:
